Ideas for using Runge--Kutta methods to preserve positivity and similar properties

### 2019-06-01 ideas.tex
Add convergence related to MRK4 with smooth initial data

\begin{remark}[About the discrete maximum principle (DMP).]
  The solution $U^L$ satisfies the following local bounds:
  \begin{align}\label{local-bounds}
    U_i^{\min}:= \min_{j\in N_i} U_j^n \leq U_i^L \leq \max_{j\in N_i} U_j^n=:U_i^{\max},
  \end{align}
  where $N_i$ is the set of DOFs that define the stencil of the discretization at the $i$-th node.
  Rewrite equation \eqref{low-order} as $U_i^L=\sum_j R_{ij}U_j^n$ with
  $R_{ij}:=\delta_{ij}-\Delta t m_i^{-1} (K_{ij}+D_{ij}^L)$, where $\delta_{ij}$ is
  the Kronecker delta function. Note that by construction
  $K_{ij}+D^L_{ij}\leq 0, \forall i\neq j\implies R_{ij}\geq 0, \forall i\neq j$.
  The diagonal entries of $R$ are positive provided $\Delta t$ is small
  enough such that $1-\Delta t m_i^{-1}(K_{ii}+D_{ii}^L)>0, \forall i$. 
  Note also that, due to partition of unity and by construction of the dissipative operator $D^L$,
  $\sum_j K_{ij}+D^L_{ij}=0\implies \sum_j R_{ij}=1$. Therefore,
  \begin{align*}
    U_i^L=\sum_j R_{ij}U_j^n\leq U_j^{\max}\sum_j R_{ij} = U_j^{\max},
  \end{align*}
  and similarly for the lower bound.
\end{remark}

@@

>>>>

\begin{remark}[About the discrete maximum principle (DMP).]
  The solution $U^L$ satisfies the following local bounds:
  \begin{align}\label{local-bounds}
    U_i^{\min}:= \min_{j\in N_i} U_j^n \leq U_i^L \leq \max_{j\in N_i} U_j^n=:U_i^{\max},
  \end{align}
  where $N_i$ is the set of DOFs that define the stencil of the discretization at the $i$-th node.
  Rewrite equation \eqref{low-order} as $U_i^L=\sum_j R_{ij}U_j^n$ with
  $R_{ij}:=\delta_{ij}-\Delta t m_i^{-1} (K_{ij}+D_{ij}^L)$, where $\delta_{ij}$ is
  the Kronecker delta function. Note that by construction
  $K_{ij}+D^L_{ij}\leq 0, \forall i\neq j\implies R_{ij}\geq 0, \forall i\neq j$.
  The diagonal entries of $R$ are positive provided $\Delta t$ is small
  enough such that
  \begin{align}\label{CFL_low_order_scheme}
    1-\Delta t m_i^{-1}(K_{ii}+D_{ii}^L)>0,\quad \forall i.
  \end{align}
  Note also that, due to partition of unity and by construction of the dissipative operator $D^L$,
  $\sum_j K_{ij}+D^L_{ij}=0\implies \sum_j R_{ij}=1$. Therefore,
  \begin{align*}
    U_i^L=\sum_j R_{ij}U_j^n\leq U_j^{\max}\sum_j R_{ij} = U_j^{\max},
  \end{align*}
  and similarly for the lower bound.
\end{remark}

@@

\clearpage
\section{Numerical experiments}

{\it Advection with smooth initial data.} Consider the domain $\Omega=(0,1)$
with periodic boundary conditions. The velocity is given by $\bfv=1$ and the initial
condition by
\begin{align}\label{smooth_initial_data}
  u(\bfx,t=0)=\frac{1}{2}(\sin(2\pi x)+1), 
\end{align}
which is clearly bounded between $[0,1]$.

\subsection{Modified RK methods with flux limiting on last stage}

{\it Advection with smooth initial data.}
We consider the one-dimensional problem with initial data given by \eqref{smooth_initial_data}
and perform a convergence study with different methods.
The spatial discretization is given by second order central finite differences.
%
In all cases, we solve the problem up to $t=1$ and report the minimum and maximum values
and the error and convergence in the $L^1$-norm given by
\begin{align}
  E=h\sum_i |U_i-u_i^{\text{exact}}|.
\end{align}

%
As a reference, we solve the problem using the standard 4th order Runge-Kutta method.
We consider \eqref{MRK_last_stage} with global bounds given by $(u_i^{\min}, u_i^{\max})=(0,1), ~\forall i$
and local bounds given by \eqref{local-bounds}.
In all cases the time step is given by $\Delta t=C h/U$ with $C=0.8, U=1$ and $h=\frac{1}{N-1}$ where $N$ is the number of DOFs
reported in the results. The results are shown in table \ref{table:MRK4_limiting_last_stage}.

{\color{red} NOTE: The time step should be chosen via \eqref{CFL_low_order_scheme}.
  However, I think $\Delta t=C h/U$ is enough for 1D with uniform meshes. I will make the change in the code and run again the tests.}

\begin{table}[h]\scriptsize
\begin{center}
\begin{tabular}{|c||c|c|c|c||c|c|c|c||c|c|c|c||} \cline{1-13}
\multicolumn{1}{|c||}{} &
\multicolumn{4}{|c||}{Standard RK4} &
\multicolumn{4}{|c||}{Modified RK4 with global bounds} &
\multicolumn{4}{|c||}{Modified RK4 with local bounds} \\ \hline
$N$ & $ \min$ &  $1-\max$ & $ E $ &  & $\min$ &  $\max$ & $E$ &   & $\min$ & $\max$ & $E$ &  \\ \hline
21  & 3.52E-2 & -1.46E-2 & 1.26E-2 & --   & 3.79E-2 & 0.9996 & 1.17E-2 & --   & 4.30E-2 & 0.9745 & 1.04E-2 & --    \\ \hline
41  & -1.58E-2 & -7.66E-3 & 2.90E-3 & 2.12 & 0       & 0.9986 & 2.64E-3 & 2.14 & 1.13E-2 & 0.9905 & 2.58E-3 & 2.01 \\ \hline
81  & -2.61E-3 & -2.88E-3 & 6.74E-4 & 2.10 & 7.03E-4 & 1      & 6.44E-4 & 2.03 & 5.38E-3 & 0.9966 & 6.42E-4 & 2.00 \\ \hline
161 & -7.78E-4 & -7.89E-4 & 1.61E-4 & 2.05 & 0       & 1      & 1.59E-4 & 2.01 & 2.02E-3 & 0.9987 & 1.59E-4 & 2.00 \\ \hline
321 & -3.39E-4 & -2.87E-4 & 3.98E-5 & 2.02 & 0       & 1      & 3.95E-5 & 2.01 & 7.82E-4 & 0.9995 & 3.96E-5 & 2.01 \\ \hline
641 & -1.30E-4 & -3.56E-5 & 9.85E-6 & 2.01 & 0       & 1      & 9.83E-6 & 2.00 & 3.11E-4 & 0.9998 & 9.85E-6 & 2.00 \\ \hline
\end{tabular}
\caption{Grid convergence with second order central finite differences in space.
  The time discretization is done via the standard RK4
  and the modified RK4 \eqref{MRK_last_stage} using global and local bounds.}
\label{table:MRK4_limiting_last_stage}
\end{center}
\end{table}
